---
title: Exploratory Data Analysis on Examining the Impact of Environmental and Demographic
  Factors on Marathon Performance
author: "Diahmin Hawkins"
date: "October 6, 2024"
output: pdf_document
editor: visual
format:
  pdf:
    documentclass: article
    geometry: margin= .5in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      echo = FALSE,
                      fig.align = "center")




#Data
library(readr)
marathon_dates <- read_csv("marathon_dates.csv")
course_record <- read_csv("course_record.csv")
aqi_values <- read_csv("aqi_values.csv")
project1 <- read_csv("project1.csv")
```

```{r}
#Packages
library(lubridate)
library(tidyverse)
library(tidyr)
library(dplyr)
library(naniar)
library(visdat)
library(kableExtra)
library(knitr)
library(gridExtra)
library(ggridges)
library(gt)
library(ggwordcloud)
library(ggplot2)
library(magick)  
library(ggplot2)
library(corrplot) 
library(reshape2)
```

```{r}



# Path to your image
fig_path <- "/Users/diahminhawkins/Documents/GitHub/Project1/weather.png"

# Load the image using magick
img <- image_read(fig_path)

# Convert image to raster for use in ggplot
img_raster <- as.raster(img)


# Example data
words <- c("Boston", "New York City", "Minneapolis", "Grandma's", "Chicago", 
           "Race", "Age", "Gender", "Weather", "Performance", 
           "Wet Bulb Globe Temperature", "Humidity")

frequencies <- c(1, 1, 1, 1, 1, 1, 5, 1, 4, 15, 3, 14)

new_frame <- data.frame(words, frequencies)

# Generate the word cloud on top of the image background
ggplot(new_frame, aes(label = words, size = frequencies)) +
  # Add the image background
  annotation_raster(img_raster, xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) +
  
  # Generate the word cloud
  geom_text_wordcloud(aes(color = frequencies)) +
  scale_size_area(max_size = 10) +
  
  # Customize the colors of the words
  scale_color_gradient(low = "yellow", high = "red") +
  
  # Remove axis titles and labels since we want the word cloud only
  theme_void()


```


```{r}
#Course Record  Data Management
course_record<- course_record%>%
  mutate(Race_Seconds= as.numeric(hms(course_record$CR)))

#Change column names from Sex... to Gender to match project1 dataset
colnames(course_record)[colnames(course_record) == "Sex (0=F, 1=M)"] ="Gender"


# Change the Race variable to a character variable
course_record <- course_record %>%
  mutate(Race = as.character(Race))



```

```{r}
# Project 1 Data Management
#Change colnames for a more readable and understanding approach
colnames(project1)[colnames(project1) == "Sex (0=F, 1=M)"] ="Gender"
colnames(project1)[colnames(project1) == "Race (0=Boston, 1=Chicago, 2=NYC, 3=TC, 4=D)"] ="Race"
colnames(project1)[colnames(project1) == "Td, C"] ="Dry bulb Temp C"
colnames(project1)[colnames(project1) == "Tw, C"] ="Wet bulb Temp C"
colnames(project1)[colnames(project1) == "%rh"] ="Percent Relative Humidity"
colnames(project1)[colnames(project1) == "Tg, C"] ="Black Globe Temp C"
colnames(project1)[colnames(project1) == "SR W/m2"] ="Solar Radiation"
colnames(project1)[colnames(project1) == "DP"] ="Dew Point in C"
colnames(project1)[colnames(project1) == "Wind"] ="Wind Speed"
colnames(project1)[colnames(project1) == "WBGT"] ="Wet Bulb Globe Temp"
colnames(project1)[colnames(project1) == "%CR"] ="Percent CR"
#change gender if 0 to F to represent female
project1$Gender<-ifelse(project1$Gender== "0","F","M")  #change gender if 0 to F to represent female else M


# Mutate the Race names from numbers to the marathon cities for better understanding
project1 <- project1 %>%
  mutate(Race = case_when(
    Race == "0" ~ "B",
    Race == "1" ~ "C",
    Race == "2" ~ "NY",
    Race == "3" ~ "TC",
    Race == "4" ~ "D"))


#Change the Race variable to a character variable
project1 <- project1 %>%
  mutate(Race = as.character(Race))

# Merge the two dataframes 
course_record_project1<-left_join(project1,course_record,
                                  by= c("Race","Gender", "Year"))

# Change the Course Percentage %CR into course seconds
course_record_project1 <- course_record_project1 %>%
  mutate(Runtimes = Race_Seconds * (1 + (`Percent CR` / 100)))
```

## Missing Data Attributes
Possibly Meet Completely at random because the the course_record wasn't dependent on the project 1 data. They carried some of the same attributes and shared covariates the we can't say one is dependent of the other.


```{r}
#Get the sum on NA's to get assumptions  (MCAR) 
sum_of_na<-sum(is.na(course_record_project1))
```

```{r, eval= FALSE}
#Examine the data
course_record_project1%>% vis_dat()
vis_miss(course_record_project1)
course_record_project1%>% glimpse()
```

```{r}

#Calculate all the missing data
sum(is.na(course_record_project1))

#Get all the missing data from each column
Missing_Data<- sapply(course_record_project1, function(x) sum(is.na(x)))
# Convert to dataframe
Missing_Data_df <- data.frame(ColumnName = names(Missing_Data), `Missing Data` = Missing_Data)

# Set names for the dataframe columns if necessary
names(Missing_Data_df) <- c("Variables", "Missing Data")

#Create variable table dataframe with description of the Marathon Dat
Variables_table<- data_frame(
  Variables= c("Race", "Year", "Gender", "Flag", "Age (yr)",
               "Percent CR", "Dry bulb Temp C","Wet bulb Temp C",        
               "Percent Relative Humidity", "Black Globe Temp C","Solar Radiation",  
               "Dew Point in C", "Wind Speed" ,"Wet Bulb Globe Temp","CR",
               "Race_Seconds","Percent_CRseconds"),
  Type= c("Character", "Numeric", "Character", "Character","Numeric",
          "Numeric","Numeric", "Numeric", "Numeric","Numeric","Numeric","Numeric",
          "Numeric","Numeric", "HMS/Numeric", "Numeric","Numeric"),
  Description= c("Race represents the marathons the participants competed, including the B=Boston Marathon,
                 C= Chicago Marathon, NY= New York City Marathon,T= Twin Cities Marathon (Minneapolis,MN),
                 D= Grandma's Marathon (Duluth, MN).",
                 "Years represented in the dataset ranging from 1993-2016.",
                 "Gender is represented by F= Female and M= Male.",
                 "Flag WBGT Thresholds. White= WBGT < 10C, Green= WBGT 10-18C, Yellow=WBGT >18-23C,
                 Red= WBGT >23-28C, and Black= WBGT > 28C",
                 "Age (yr) represents the ages of the participants.",
                 "Percent CR is the percent off current course record for gender.",
                 "Dry bulb Temp Celcius is the air temperature without taking into account of the humidity or any
                 moisture.",
                 "Wet bulb Temp Celcius is a measure of temperature that reflects both the heat and humidity in the air. Wet bulb temperature gives you an idea of how temperature feels when you take humidity into account. ",
                 "Percent Relative Humidity how much moisture is in the air compared to the maximum amount of moisture the air can hold at a given temperature. Gives an idea of how humid it feels outside.",
                 "Black Globe Temp Celcius  indicates how hot it feels in direct sunlight.It considers temperature, humidity, wind speed, sun angle, and cloud cover to provide a holistic view of the stress placed on the body in hot environments.",
                 "Solar Radiation in Watts per meter squared is the energy emitted by the sun, which travels through space and reaches the Earth as light and heat.",
                 "Dew Point in Celcius is the temperature the air needs to be cooled to (at constant pressure)                  in order to achieve a relative humidity (RH) of 100%. At this point the air cannot hold more                   water in the gas form. If the air were to be cooled even more, water vapor would have to come out of the atmosphere in the liquid form, usually as fog or precipitation.",
                 "Wind Speed in Km/hr.",
                 "Wet Bulb Globe Temp measures the combined effect of temperature, humidity, wind speed, and solar radiation on humans. Formula WBGT= 0.7 x Tw + 0.2 x Tg+ 0.1 xTd.",
                 "CR is the course record for each marathon.",
                 "Race_Seconds is the course record  measured in seconds.",
                 "Percent_CRseconds is the converted gender percentage into seconds."
                 )
)

Missing_Data_df$Variables <- as.character(Missing_Data_df$Variables)
Variables_table$Variables <- as.character(Variables_table$Variables)
merged_df <- merge(Missing_Data_df, Variables_table, by = "Variables", all = TRUE)


# Create the table with kable and customize with kableExtra
kable(merged_df, "latex", booktabs = TRUE, caption = "Marathon Runners' Data Description") %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  column_spec(1, width = "3cm") %>%
  column_spec(2, width = "2cm") %>%
  column_spec(3, width = "2cm") %>%
  column_spec(4, width = "8cm")


course_record_project1<-na.omit(course_record_project1)

```



#AIM 1: Examine effects of increasing age on marathon performance in men and women
In the course record data we have Pecent CR which give percentages by gender
In the project 1 data we have course record overall has race seconds



```{r}
#Find the minimum age 
minimimum_age<-min(course_record_project1$`Age (yr)`)

#Find the maximum age
maximum_age<-max(course_record_project1$`Age (yr)`)

#Create Age Ranges/ Age Breaks to Categorize the groups
course_record_project1$age_ranges<- cut(
  course_record_project1$`Age (yr)`,
  breaks = c(0, 25, 35, 45, 55, 65, 75, Inf),  # Custom breaks for the new age ranges
  labels = c("<15-25", "26-35", "36-45", "46-55", "56-65", "66-75", "76+")

)

#Get Counts By Age Group to see balance
age_range_counts <- course_record_project1 %>%
  group_by(age_ranges)%>%
  summarise(count=n())


#Marathon Performance by age by Race 
marathon_performance_by_age<- course_record_project1%>%
  select(Race,Year, Gender, age_ranges, Runtimes, `Age (yr)`)%>%
  group_by(Race, Gender, age_ranges)




```

```{r, eval= FALSE}
#Get the best course_record from each race 
best_course_race<- course_record_project1 %>%
  filter(Runtimes <= Race_Seconds)%>%
  group_by(Race, Gender, age_ranges)%>%
  summarise(count=n())


# Rename some of the column names
best_course_race <- best_course_race %>%
  rename(
    `Marathon` = Race,          # Rename 'Race' to 'Race Name'
    `Gender` = Gender,           # Keep the 'Gender' column as is (optional)
    `Age Range` = age_ranges,    # Rename 'age_ranges' to 'Age Range'
    `Number of Participants` = count  # Rename 'count' to 'Number of Participants'
  )
# Create the table with the new column names and specified styling
best_course_race %>%
  kbl(caption = "<div style='text-align:center; font-size:24px; font-weight:bold;'>Marathon Runners with High Performance by Race</div>") %>%
  kable_classic(full_width = F, html_font = "Cambria", font_size = 20) %>%
  kable_styling(position = "center", font_size = 16) 


worst_course_race <- course_record_project1 %>%
  filter(Runtimes >= Race_Seconds) %>%
  group_by(Race,Gender, age_ranges)%>%
  summarise(count=n())



just_gender_age_ranges<- course_record_project1 %>%
  filter(Runtimes >= Race_Seconds) %>%
  group_by(Gender, age_ranges)%>%
  summarise(count=n())

worst_course_race %>%
  kbl(caption = "Number of Marathon Runners that Did not beat the Course Record by Race, Gender, and Age Group") %>%
  kable_classic(full_width = F, html_font = "Times New Roman", font_size= 20)
  

# Create bar plot od the Worst Course Race varaiable for easier read
ggplot(worst_course_race, aes(x = Race, y = count, fill = Gender)) +
  geom_bar(stat = "identity", position = "dodge") +  # Create a bar plot, grouped by Gender
  labs(title = "Number of Marathon Runners that Did not beat the Course Record by Race, Gender, and Age Group",
       x = "Race",
       y = "Marathon Runners") +
  scale_fill_manual(values = c("F" = "hotpink", "M" = "royalblue")) +  # colors pink for F and blue for M
  facet_wrap(~ age_ranges) + 
  theme_minimal()  


```

```{r, fig.width=10, fig.height=6}
# Line plot of the Marathon Runners Performance by Age 
age_plot<-ggplot(marathon_performance_by_age, aes(x = `Age (yr)`, y = Runtimes, color = Gender)) +
  geom_point(alpha = 0.05) +
  geom_smooth(se = FALSE, linewidth = 1.5) +
  labs(title = "Effects of Age on Marathon Performance in Men and Women",
       x = "Ages",
       y = " Runtimes (Performance)") +
  scale_color_manual(values = c("F" = "hotpink", "M" = "royalblue")) +
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5) )



age_plot


# Create the boxplot to visualize the grouping 
age_boxplot<-ggplot(marathon_performance_by_age, aes(x = age_ranges, y =Runtimes, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Effects of Age on Marathon Performance in Men and Women",
       x = "Age Ranges",
       y = "Runtimes (Performance)") +
   scale_fill_manual(values = c("F" = "hotpink", "M" = "royalblue"))+
  theme_classic()


age_boxplot

# Create boxplot stratified by diffferent Races 
ggplot(marathon_performance_by_age, aes(x = age_ranges, y = Runtimes, fill = Gender)) +
  geom_boxplot() +
  facet_wrap(~ Race) +  # Facet by Race to create separate plots for each race
  labs(title = "Effects of Marathon Performance  Stratified by Race, Gender, and Age Group",
       x = "Age Ranges",
       y = "Runtimes (Performance)") +
  theme_classic()+
  scale_fill_manual(values = c("F" = "hotpink", "M" = "royalblue"))+
  theme(axis.text.x= element_text(angle =45, vjust= 1, hjust = 1),
        plot.title = element_text(hjust = 0.5) )


marathon_performance_summary_table <-marathon_performance_by_age%>%
  group_by(age_ranges) %>%
  summarise(
    Count=n(),
    Mean= mean(Runtimes),
    Q1_Runtime = quantile(Runtimes, 0.25, na.rm = TRUE),
    Median_Runtime = median(Runtimes, na.rm = TRUE),       # Median of Runtimes
     # Lower quartile (25th percentile)
    Q3_Runtime = quantile(Runtimes, 0.75, na.rm = TRUE),    # Upper quartile (75th percentile)
    IQR_Runtime = IQR(Runtimes, na.rm = TRUE) # IQR of Runtimes
  )

marathon_performance_summary_table %>%
  gt() %>%
  tab_header(
    title = "Marathon Performance Summary by Age Range"
  ) %>%
  cols_label(
    age_ranges = "Age Ranges",
    Count = "Marathon Runners",
    Mean = "Mean Runtimes",
     Q1_Runtime = "1Q",
    Median_Runtime = "Median Runtimes",
    Q3_Runtime = "3Q",
    IQR_Runtime = "IQR"
  ) %>%
  fmt_number(
    columns = vars(Mean, Median_Runtime, Q1_Runtime, Q3_Runtime, IQR_Runtime),
    decimals = 0 # Set decimal places for summary statistics
  )
```






##  Explore the impact of environmental conditions on marathon performance, and whether the impact differs across age and gender.
Higher temperatures lead to slower times 
stratify by gender age 


```{r}
 # Create Dataframe to allocate for other environmental conditions
 environmental_conditions <- course_record_project1 %>%
  select(Race, Gender, `Age (yr)`, Runtimes, `Dry bulb Temp C`, `Wet bulb Temp C`,
  `Percent Relative Humidity`, `Black Globe Temp C`, `Solar Radiation`,`Dew Point in C`, 
  `Wind Speed`, `Wet Bulb Globe Temp`, age_ranges) %>%   
  group_by( Gender, `Age (yr)`)

```


```{r, fig.width=10, fig.height=6}
#Observing the linear model approach to see the impact on performance due to weather conditions and
# observer the difference in gender and age 
# Linear to evaluate statistical significance
model <- lm(Runtimes ~ `Dry bulb Temp C` + `Percent Relative Humidity` + `Black Globe Temp C` + 
         `Wind Speed` + `Dew Point in C` + `Solar Radiation` + `Age (yr)` + `Wet Bulb Globe Temp` + Gender +
         `Dry bulb Temp C`:`Age (yr)` + `Percent Relative Humidity`:`Age (yr)` + 
         `Black Globe Temp C`:`Age (yr)` +`Wind Speed`:`Age (yr)` + `Dew Point in C`:`Age (yr)` + 
         `Solar Radiation`:`Age (yr)` + `Wet Bulb Globe Temp`:`Age (yr)` + Gender:`Age (yr)` +
         `Dry bulb Temp C`:Gender + `Percent Relative Humidity`:Gender + `Black Globe Temp C`:Gender +
          `Wind Speed`:Gender + `Dew Point in C`:Gender + `Solar Radiation`:Gender +
          `Wet Bulb Globe Temp`:Gender, data = environmental_conditions)

model_table <- round(summary(model)$coefficients, 5)

# Use kable to create the summary table
model_table %>%
  kbl(caption = "Impact of Environmental Conditions on Marathon Performance by Age and Gender",
      booktabs = TRUE, escape = FALSE, align = "c") %>%
  kable_styling(full_width = FALSE, latex_options = c('hold_position'))%>%
  column_spec(1, bold = TRUE, color="black", border_right = TRUE)%>%
  row_spec(1, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(2, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(3, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(4, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(6, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(7, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(8, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(9, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(10, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(11, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(12, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(13, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(15, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(16, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(17, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(18, bold = TRUE, color="black", background = "#87CEEB")%>%
  row_spec(20, bold = TRUE, color="black", background = "#87CEEB")
  

#Create Correlation plot
# Observing  and Including all the numeric variables
numeric_data <- course_record_project1%>%
  select(Runtimes, `Dry bulb Temp C` , `Percent Relative Humidity` , `Black Globe Temp C` ,
       `Wind Speed` , `Dew Point in C` ,`Solar Radiation` , `Age (yr)` ,`Wet Bulb Globe Temp`)

# Compute the correlation matrix using complete observations
cor_matrix <- cor(numeric_data, use = "complete.obs")  # Use complete.obs to ignore NAs

# Melt the correlation matrix for ggplot2
cor_data <- melt(cor_matrix)

#Environmental conditions correlation plot 
environmental_conditions_plot<-ggplot(data = cor_data, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  scale_fill_gradient2(low = "hotpink", high = "royalblue", mid = "white",
                       midpoint = 0, limit = c(-1, 1), space = "Lab",
                       name = "Correlation") +
  labs(title = "Correlation of Environmental Conditions on Marathon Performance") +
  theme_minimal(base_family = "Times") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        legend.position = "bottom")
environmental_conditions_plot
  


# Dry bulb plot
 dry_bulb_plot<-ggplot(environmental_conditions, aes(x = `Dry bulb Temp C`, y = Runtimes, color = Gender)) +
  geom_point(alpha=.1) +  # Scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "black") +  # Linear regression line with confidence interval
  facet_wrap(~ age_ranges) +  # Facet the plot by age_ranges
  theme_classic() + 
  labs(title = "Marathon Performance vs. Dry Bulb Temperature by Age Ranges",
       x = "Dry Bulb Temperature (째C)",
       y = "Marathon Performance (Percent CR Seconds)")+
  scale_color_manual(values = c("M" = "royalblue", "F" = "hotpink"))

 
  dry_bulb_plot
  
  
# Humidity Percentages Plot
relative_percent_humidity<- ggplot(environmental_conditions, aes(x = `Percent Relative Humidity`, 
                            y = Runtimes, color = Gender)) +
                            geom_point(alpha=.1) +  # Scatter plot
                            geom_smooth(method = "lm", formula = y ~ x, color = "black") +  
                            # Linear regression line with confidence interval
                            facet_wrap(~ age_ranges) +  # Facet the plot by age_ranges
                            theme_classic() +  
                            labs(title = "Marathon Performance vs. Relative Humidity Stratified by Age Ranges ",
                                    x = "Pecent Relative Humidity %",
                                    y = "Marathon Performance (Runtimes")+
                            scale_color_manual(values = c("M" = "royalblue", "F" = "hotpink"))

relative_percent_humidity

 # Black Globe Temperature
black_globe_temp_graph<- ggplot(environmental_conditions, aes(x = `Black Globe Temp C`, 
                          y = Runtimes, color = Gender)) +
                          geom_point(alpha=.1) +  # Scatter plot
                          geom_smooth(method = "lm", formula = y ~ x, color = "black") +  
                          # Linear regression line with confidence interval
                          facet_wrap(~ age_ranges) +  # Facet the plot by age_ranges
                          theme_classic() + 
                          labs(title = "Marathon Performance vs. Black Globe Temperature Stratified by Age Groups",
                          x = "Black Globe Temp  (째C)",
                          y = "Marathon Performance (Runtimes)")+
                          scale_color_manual(values = c("M" = "royalblue", "F" = "hotpink"))

black_globe_temp_graph


 # Wet Bulb Temperature
 wet_bulb_graph<-ggplot(environmental_conditions, aes(x = `Wet Bulb Globe Temp`, 
                       y = Runtimes, color = Gender)) +
                       geom_point(alpha=.1) +  
                       geom_smooth(method = "lm", formula = y ~ x, color = "black") +  
                       # Linear regression line with confidence interval
                       facet_wrap(~ age_ranges) +  # Facet the plot by age_ranges
                       theme_classic() +  
  labs(title = "Marathon Performance vs. Wet Bulb Globe Temp Stratified by Age Groups",
       x = "Wet Bulb Globe Temp (째C)",
       y = "Marathon Performance (Runtimes)")+
  scale_color_manual(values = c("M" = "royalblue", "F" = "hotpink"))

 wet_bulb_graph
 
 #Solar Radiation Graph
 solar_radiation_graph<-ggplot(environmental_conditions, aes(x = `Solar Radiation`, 
                    y = Runtimes, color = Gender)) +
                    geom_point(alpha=.1) +  # Scatter plot
                    geom_smooth(method = "lm", formula = y ~ x, color = "black") +  
                   # Linear regression line with confidence interval
                    facet_wrap(~ age_ranges) +  # Facet the plot by age_ranges
                    theme_classic() + 
                    labs(title = "Marathon Performance vs. Solar Radiation Stratified by Age Groups",
                    x = "Solar Radiation",
                    y = "Marathon Performance (Runtimes)")+
                    scale_color_manual(values = c("M" = "royalblue", "F" = "hotpink"))
 
solar_radiation_graph

 
black_globe_temp_graph <- ggplot(environmental_conditions, aes(x = `Black Globe Temp C`, y = Runtimes, color = Gender)) +
  geom_point(alpha=.1) +  # Scatter plot
  geom_smooth(method = "lm", formula = y ~ x, color = "black") +  # Linear regression line with confidence interval
  facet_wrap(~ age_ranges) +  # Facet the plot by age_ranges
  theme_classic() + 
  labs(title = "Marathon Performance vs. Black Globe Temperature Stratified by Age Groups",
       x = "Black Globe Temp  (째C)",
       y = "Marathon Performance (Runtimes)") +
    # Use scale_color_manual to set custom colors for Male and Female
  scale_color_manual(values = c("M" = "royalblue", "F" = "hotpink"))
 
black_globe_temp_graph 

```

## Aim 3: Identify the weather parameters (WBGT, Flag conditions, temperature, etc) that have the largest impact on marathon performance.


```{r}
#Data Management of the marathon data set and air quality 
#Change Race Names
marathon_dates <- marathon_dates %>%
   mutate(marathon = case_when(
     marathon == "NYC" ~ "NY",              
     marathon == "Grandmas" ~ "D",
     marathon == "Boston" ~ "B",
     marathon == "Twin Cities" ~"TC" 
  ))

#Change column names to match course_record_project column names
colnames(marathon_dates)[colnames(marathon_dates) == "marathon"] ="Race"
colnames(marathon_dates)[colnames(marathon_dates) == "year"] ="Year"

# Change formatting  of the dates
marathon_dates <- marathon_dates %>%
 mutate(date = as.Date(date, format = "%Y-%m-%d"))

# Combine the marathon dates datframe to my current dataframe by using left_join
course_record_project1 <- course_record_project1%>%
    left_join(marathon_dates, by = c("Race", "Year"))


# Change the marathon variable to Race to match corresponding data and change race names
aqi_values <- aqi_values %>%
   rename(Race = marathon) %>%
     mutate(
     Race = case_when(
     Race == "NYC" ~ "NY",              
     Race == "Grandmas" ~ "D",
     Race == "Boston" ~ "B",
     Race == "Twin Cities" ~"TC" 
  ),
    date = as.Date(date_local, format = "%Y-%m-%d"),
    Year = as.numeric(format(date, "%Y"))
      ) %>%
   select(-date_local) #Remove the date_local variable


# calculate average ozone ppm (8-hour avg)
avg_ppm <- aqi_values %>%
   filter(units_of_measure == "Parts per million",
          sample_duration == "8-HR RUN AVG BEGIN HOUR") %>%
          group_by(Race, Year, date) %>%
          summarize(avg_ppm = mean(arithmetic_mean, na.rm = T)) %>%
          ungroup()



# Merge data_frame to current dataframe
 course_record_project1 <-  course_record_project1 %>%
   left_join(avg_ppm, by = c("Race", "Year", "date"))
 
 
```





```{r}
weather_parameters<-course_record_project1%>%
  select(Gender, `Age (yr)`, Runtimes, `Dry bulb Temp C`, `Wet bulb Temp C`,
  `Percent Relative Humidity`, `Black Globe Temp C`, `Solar Radiation`,`Dew Point in C`, 
  `Wind Speed`, `Wet Bulb Globe Temp`, age_ranges, Flag)
```

```{r}


# Create the box plot with custom Flag colors and filled boxplots
flag_exam <- ggplot(weather_parameters, aes(x = Gender, y = Runtimes, fill = Flag)) +
  geom_boxplot() +  # Use geom_boxplot for creating a boxplot
  theme_bw() +      # Black and white theme for a clean plot
  scale_fill_manual(values = c(
    "White" = "white",   # Replace these with the Flag categories and the colors you want to assign
    "Green" = "green",
    "Black" = "black",
    "Yellow" = "yellow",  # Add more colors for other Flag categories as needed
    "Red" = "red"
  )) +
  labs(
    title = "Marathon Performance by Flag Conditions Stratified by Gender",
    x = "Gender",
    y = "Marathon Performance (Runtimes)",
    fill = "Flag Color"  # Title for the legend 
  ) +
  theme((legend.position = "right"),
    plot.title = element_text(hjust = 0.5, size = 12)) # Center and set title size to 12


flag_exam


flag_age_ranges<- ggplot(weather_parameters, aes(x = age_ranges, y = Runtimes, fill = Flag)) +
  geom_boxplot() +  # Use geom_boxplot for creating a boxplot
  theme_bw() +      # Black and white theme for a clean plot
  scale_fill_manual(values = c(
    "White" = "white",   # Replace these with the Flag categories and the colors you want to assign
    "Green" = "green",
    "Black" = "black",
    "Yellow" = "yellow",  # Add more colors for other Flag categories as needed
    "Red" = "red"
  )) +
  labs(
    title = "Marathon Performance by Flag Conditions Stratified by Age Ranges",
    x = "Age Ranges",
    y = "Marathon Performance (Runtimes)",
    fill = "Flag Color"  # Title for the legend
  ) +
  theme(
    (legend.position = "right"),
    plot.title = element_text(hjust = 0.5, size = 12)) 
   
flag_age_ranges













```

```{r, fig.width=10, fig.height=6}


# Select only numeric columns from the data frame for correlation plot
airquality<- course_record_project1%>%
  select(Runtimes, `Dry bulb Temp C` , `Percent Relative Humidity` , `Black Globe Temp C` ,
       `Wind Speed` , `Dew Point in C` ,`Solar Radiation` , `Age (yr)` ,`Wet Bulb Globe Temp` , Year, 
       avg_ppm)


#Compute the correlation matrix
cor_matrix2 <- cor(airquality, use = "complete.obs")  # Use complete.obs to ignore NAs

# Melt the correlation matrix for ggplot2
library(reshape2)
cor_data2<- melt(cor_matrix2)


# Visualize the Correlation Plot of the the Weather Parameters
ggplot(data=cor_data2, aes(x= Var1, y= Var2, fill=value))+
  geom_tile(color= "white")+
  geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
  scale_fill_gradient2(low= "hotpink", high="royalblue", mid = "white",
  midpoint=0, 
  limit=c(-1,1), 
  space="Lab",
  name="Correlation")+
  labs(title="Weather Parameters with the Largest Impact on Marathon Performance Correlation Plot")+
  theme_minimal(base_family="Times")+
  theme(axis.text.x= element_text(angle =45, vjust= 1, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 12))  # Center and set title size to 20
  
  

# Summarize data by Flag to calculate the median, IQR, Q1 (25th percentile), and Q3 (75th percentile) for Runtimes
flag_summary <- course_record_project1 %>%
  group_by(Flag,age_ranges) %>%
  summarise(
    Median_Runtime = median(Runtimes, na.rm = TRUE),       # Median of Runtimes
    IQR_Runtime = IQR(Runtimes, na.rm = TRUE),             # IQR of Runtimes
    Q1_Runtime = quantile(Runtimes, 0.25, na.rm = TRUE),   # Lower quartile (25th percentile)
    Q3_Runtime = quantile(Runtimes, 0.75, na.rm = TRUE)    # Upper quartile (75th percentile)
  )





```

































 


Code Appendix
```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
